Working document to try to explain why/how ATAs influence coexistence in an intuitive way for referee 3

### 2023-07-28 paper/ExplanationOfATAEffectsWorkingDoc.tex
explanatory text pretty developed at this point, though I will edit, and may add another example using the lognormal fecundities lottery model

If temporal niche differentiation occurs but is partial, 
then $E_i$ and $C_{i \bs i}$ will be positively but imperfectly correlated with
each other, corresponding to similar but non-identical responses of species to
environmental fluctuation (see SI section \ref{SIsect:temporal_niche_diff}); 
but how do ATA effects then happen? We here extend the intuition
introduced in the Discussion and provide examples.

Storage can occur if population
increases during good years outweigh decreases during bad years. For instance,
if $r_i(E_i,C_{i \bs i})$ is very large when the balance of $E_i$ and 
$C_{i \bs i}$ is favorable, but only moderately small when this balance is
unfavorable, the expected value $\E[r_i(E_i,C_{i \bs i})]$ in (\ref{eq:expri})
will be positive. But the potential for such storage of population growth
may differ between when $E_i$ and $C_{i \bs i}$ are both high compared
to when they are both low. 
For instance, for the beta fecundities lottery model, $E_i/C_{i \bs i}$, a quantity
which features prominently in $r_i(E_i,C_{i \bs i})=\log(1-\delta+E_i/C_{i \bs i})$,
takes values bounded above by $\frac{2\eta_i \delta}{\eta_j}$ 
and bounded below by $\frac{\eta_i \delta}{2 \eta_j}$ when $E_i$ and $C_{i \bs i}$
are both in the upper halves of their assumed beta distribtions (recall that 
these quantities were assumed to be distributed, respectively, as $\eta_1 \beta$
and $\eta_2 \beta/\delta$ for $\beta$ a beta-distributed random variable with parameters 
$0.5$ and $0.5$). These bounds are separated only by a factor of four. 
On the other hand, when $E_i$ and $C_{i \bs i}$ are both in the lower halves 
of their distributions, $E_i/C_{i \bs i}$ can take any positive value because
then either $E_i$ or $C_{i \bs i}$ can take values arbitrarily close to $0$. 
Because values of $E_i/C_{i \bs i}$ have the potential to be much larger when 
$E_i$ and $C_{i \bs i}$ are both in the lower halves of their distributions
than when these quantities are both in their upper halves, and because
the growth rate, $\log(1-\delta+E_i/C_{i \bs i})$, buffers low values of 
$E_i/C_{i \bs i}$ (the infemum of the growth rate is $\log(1-\delta)$, whereas the supremum is
infinite), the storage effects contributed by the left tails of 
$E_i$ and $C_{i \bs i}$ have the potential to be much greater than those contributed
by the right tails. Essentially, values of $\log(1-\delta+E_i/C_{i \bs i})$
occurring when $E_i$ and $C_{i \bs i}$ are both in the upper halves of their
distributions can never exceed $\log(1-\delta+\frac{2\eta_i \delta}{\eta_j})$,
whereas values ocurring when $E_i$ and $C_{i \bs i}$ are both in the lower
halves of their distributions can be arbitrarily large (though with decreasing 
probability for larger values). Apparently, the years with the most potential to 
contribute to storage
effects for the beta-fecundities lottery model are those for which the environment
is actually poorer than average, because, for those years, competition can be very low,
so that the ratio $E/C$ is large. Years for which the
environment is exceptionally good actually cannot contribute much to storage effects for 
this model, because for those years, $E/C$ cannot be large.
It is the imbalance between $E$ and $C$ that allows for storage effects. 
Counterintuitively, for this model, that imbalance is greatest when the environment is 
poor. This result is counterintuitive because one may be used to thinking of storage effects
occuring when reproductive pulses from environmentally good years sustain the population
through environmentally poor years. But, in fact, good years for population growth may
happen principally when the environment is poor, if those are also the times when 
competition is very low, as for this model.

Fig. \ref{fig:beta_fecundidities_explanatory} provides quantitative backing to
the ideas of the prior paragraph. The growth function $\ln(1-\delta+E/C)$ that pertains 
to the lottery model
is plotted with colors and contours on Fig. \ref{fig:beta_fecundidities_explanatory}A-C.
Samples were taken from the distribution $(E_i,C_{i \bs i})$ for the beta 
fecundities lottery model for left- and right-tailed 
associations between $E_i$ and $C_{i \bs i}$, and are plotted as
points on Fig. \ref{fig:beta_fecundidities_explanatory}A and C, respectively. 
Samples were also taken from the distribution $(E_i^\nat,C_{i \bs i}^\nat)$
(which is the same regardless of whether the original $(E_i,C_{i \bs i})$
distribution showed left- or right-tail association) and are plotted on 
Fig. \ref{fig:beta_fecundidities_explanatory}B. 
The dashed lines labelled $L_2$ on A-C separate the upper-right $50\%$
of each distribution from the lower-left $50\%$. The dashed lines labeled
$L_1$ separate the upper-left $50\%$ from the lower-right $50\%$.
The quantity $\E \left[ \ln \left( 1-\delta+\frac{E_i}{C_{i \bs i}} \right) \right]$ can be
well approximated for left- and right-tail associations by averaging function 
values, depicted with color and contours, over the points on panels A and C, respectively. The quantity
$\E \left[ \ln \left( 1-\delta+\frac{E_i^{||}}{C_{i \bs i}^{||}} \right) \right]$
can be well approximated by averaging function 
values over the points on B. So ATA effects for the model with left-tail association
between $E_i$ and $C_{i \bs i}$ (respectively, right-tail association) are given by 
computing the average function value over points on panel A (respectively, C)
and then subtracting the average function value over points on panel B.
Contributions to the storage effect of points in the left portions of 
the $E,C$ distribution (the region labeled ``left'' on panels A and C)
can be accurately estimated by averaging function values across
points in that region; likewise storage contributions of points in the 
``right'' region can be estimated by averaging function values over 
points in that region.

The details of the averaging process described above are pictured in 
Fig. \ref{fig:beta_fecundidities_explanatory}, panels D-F;
we now describe how. 
Signed distances of points on Fig. \ref{fig:beta_fecundidities_explanatory} 
A-C from the lines $L_1$ correspond to 
the quantities $\log_{10}(E_1/C_{1 \bs 1})$ on A, C and to 
$\log_{10}(E_1^\nat/C_{1 \bs 1}^\nat)$ on B. The distributions of these 
quantities across points to the left of $L_2$ are pictured 
in the lower horizontal-axis historgrams on panels D-F, respectively;
and distributions for points to the right of $L_2$ are 
pictured in the upper horizontal-axis histograms. 
For the left-tailed case (A, D), distributions to the

\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/PostHocExplanatoryFig_betaFecund.jpg}
\caption{A detailed look at ATA effects for the beta fecundities lottery model.
The figure gives a pictorial representation of the calculation of  
$\E \left[ \ln \left( 1-\delta+\frac{E_i}{C_{i \bs i}} \right) \right]$ (A, C, D, F) and 
$\E \left[ \ln \left( 1-\delta+\frac{E_i^{||}}{C_{i \bs i}^{||}} \right) \right]$ (B, E)
for the model, considering both left- (A, D) and right-tail (C, F) associations
between $E_i$ and $C_{i \bs i}$.
The colors and contours on A-C $\ln(1-\delta+E/C)$, whereas the points show a sample
from the distribution $(E_i,C_{i \bs i})$ (A, C) or $(E_i^{||},C_{i \bs i}^{||})$ (B).
The quantity $\E \left[ \ln \left( 1-\delta+\frac{E_i}{C_{i \bs i}} \right) \right]$ can be
well approximated for left- and right-tail associations by averaging function 
values over the points on panels A and C, respectively. The quantity
$\E \left[ \ln \left( 1-\delta+\frac{E_i^{||}}{C_{i \bs i}^{||}} \right) \right]$
can be well approximated by averaging function 
values over the points on B. Values of $\log_{10}(E/C)$ for the points on A-C correspond 
to signed distances of the points from the upward sloping dashed lines. 
We used parameters $\eta_1 = 1$, $\eta_2 = 1.2$, and and $\delta=0.6$. }\label{fig:beta_fecundidities_explanatory}
\end{figure}

>>>>

If temporal niche differentiation occurs but is partial, 
then $E_i$ and $C_{i \bs i}$ will be positively but imperfectly correlated with
each other, corresponding to similar but non-identical responses of species to
environmental fluctuation (SI section \ref{SIsect:temporal_niche_diff}). 
But how do ATA effects then happen? We here extend the intuition
introduced in the Discussion and provide examples.

First, we illustrate using the beta fecundities lottery model that
the potential for storage to occur when 
$E$ and $C$ are both higher than average, versus the potential  
when $E$ and $C$ are
both lower than average, are not necessarily equivalent.
Storage can occur if population
increases during good years outweigh decreases during bad years. For instance,
if $r_i(E_i,C_{i \bs i})$ is very large when the balance of $E_i$ and 
$C_{i \bs i}$ is favorable, but only moderately small when this balance is
unfavorable, the expected value $\E[r_i(E_i,C_{i \bs i})]$ in (\ref{eq:expri})
will be positive. But the potential for such storage of population growth
may differ between when $E_i$ and $C_{i \bs i}$ are both high compared
to when they are both low. 
For instance, for the beta fecundities lottery model, $E_i/C_{i \bs i}$, a quantity
which features prominently in $r_i(E_i,C_{i \bs i})=\log(1-\delta+E_i/C_{i \bs i})$,
takes values bounded above by $\frac{2\eta_i \delta}{\eta_j}$ 
and bounded below by $\frac{\eta_i \delta}{2 \eta_j}$ when $E_i$ and $C_{i \bs i}$
are both in the upper halves of their assumed beta distribtions (recall that 
these quantities were assumed to be distributed, respectively, as $\eta_1 \beta$
and $\eta_2 \beta/\delta$ for $\beta$ a beta-distributed random variable with parameters 
$0.5$ and $0.5$). These bounds are separated only by a factor of four. 
On the other hand, when $E_i$ and $C_{i \bs i}$ are both in the lower halves 
of their distributions, $E_i/C_{i \bs i}$ can take any positive value because
then either $E_i$ or $C_{i \bs i}$ can take values arbitrarily close to $0$. 
Because values of $E_i/C_{i \bs i}$ have the potential to be much larger when 
$E_i$ and $C_{i \bs i}$ are both in the lower halves of their distributions
than when these quantities are both in their upper halves, and because
the growth rate, $\log(1-\delta+E_i/C_{i \bs i})$, buffers low values of 
$E_i/C_{i \bs i}$ (the infemum of the growth rate is $\log(1-\delta)$, whereas the supremum is
infinite), the storage effects contributed by the left tails of 
$E_i$ and $C_{i \bs i}$ have the potential to be much greater than those contributed
by the right tails. Essentially, annual growth-when-rare rates for species $i$
that occur when $E_i$ and $C_{i \bs i}$ are both in the upper halves of their
distributions can never exceed $\log(1-\delta+\frac{2\eta_i \delta}{\eta_j})$,
whereas rates ocurring when $E_i$ and $C_{i \bs i}$ are both in the lower
halves of their distributions can be arbitrarily large. 
Apparently, the years with the most potential to 
contribute to storage
effects for the beta-fecundities lottery model are those for which the environment
is actually poorer than average, because, for those years, competition can be very low,
so that the ratio $E_i/C_{i \bs i}$ is large. Years for which the
environment is exceptionally good actually cannot contribute much to storage effects for 
this model, because for those years, $E_i/C_{i \bs i}$ cannot be very large.
It is the \emph{imbalance} between $E$ and $C$ that allows for storage effects,
not solely the quality of environment. 
Counterintuitively, that imbalance can be greater for this model when the environment is 
sub-optimal.

Fig. \ref{fig:beta_fecundidities_explanatory} provides quantitative backing to
the ideas of the prior paragraph, so we orient the reader 
to the figure before using it below to further explain
ATAs. The growth function $\ln(1-\delta+E/C)$ that pertains to the lottery model
is plotted with colors and contours on Fig. \ref{fig:beta_fecundidities_explanatory}A-C.
Samples were taken from the distribution $(E_i,C_{i \bs i})$ for the beta 
fecundities lottery model for left- and right-tailed 
associations between $E_i$ and $C_{i \bs i}$, and are plotted as
points on Fig. \ref{fig:beta_fecundidities_explanatory}A and C, respectively. 
Samples were also taken from the distribution $(E_i^\nat,C_{i \bs i}^\nat)$
(which is the same regardless of whether the original $(E_i,C_{i \bs i})$
distribution showed left- or right-tail association) and are plotted on 
Fig. \ref{fig:beta_fecundidities_explanatory}B. 
The dashed lines labelled $L_2$ on A-C separate the upper-right $50\%$
of each distribution from the lower-left $50\%$, and those portions are labelled. 
The quantity $\E \left[ \ln \left( 1-\delta+\frac{E_i}{C_{i \bs i}} \right) \right]$ can be
well approximated, for left- and right-tail association cases, by averaging values of the
function depicted with color and contours over the points on panels A and C, 
respectively. The quantity
$\E \left[ \ln \left( 1-\delta+\frac{E_i^{||}}{C_{i \bs i}^{||}} \right) \right]$
can be well approximated by averaging function 
values over the points on B. So ATA effects for the model with left-tail association
between $E_i$ and $C_{i \bs i}$ (respectively, right-tail association) are given by 
computing the average function value over points on panel A (respectively, C)
and then subtracting the average function value over points on panel B.
Contributions to the storage effect of points in the left portions of 
the $E,C$ distribution (the region labeled ``left'' on panels A and C)
can be accurately estimated by averaging function values across
points in that region; likewise storage contributions of points in the 
``right'' region can be estimated by averaging function values over 
points in that region.

The details of the averaging processes described above are pictured in 
Fig. \ref{fig:beta_fecundidities_explanatory}, panels D-F. 
Signed distances of points on Fig. \ref{fig:beta_fecundidities_explanatory} 
A-C from the lines $L_1$ correspond to 
the quantities $\log_{10}(E_1/C_{1 \bs 1})$ on A, C and to 
$\log_{10}(E_1^\nat/C_{1 \bs 1}^\nat)$ on B. The distributions of these 
quantities across points in the ``left'' region are pictured 
in the lower horizontal-axis histograms on panels D-F, respectively;
and distributions for points in the ``right'' region are 
pictured in the upper horizontal-axis histograms. 
``Right'' histograms on the horizontal axis of D,F are always highly 
concentrated, corresponding to the fact that, for the beta fecundities
lottery model, $E_i/C_{i \bs i}$ cannot vary much when $E_i$ and 
$C_{i \bs i}$ are in the upper portions of their distributions. 
Likewise, the ``right'' histogram on E is highly concentrated because
$E_i^\nat/C_{i \bs i}^\nat$ cannot vary much when $E_i^\nat$ and 
$C_{i \bs i}^\nat$ are in the upper portions of their distributions.
In contrast, ``left'' histograms on the horizontal axis of D-F
can be highly variable, though only when $E_i$ and $C_{i \bs i}$
(or $E_i^\nat$ and $C_{i \bs i}^\nat$) are not tightly 
correlated in the left portions of their distributions.
The function pictured with a black line on panels D-F is simply the function
$z \rightarrow \ln(1-\delta+10^z)$ that takes $z=\log_{10}(E_1,C_{1 \bs 1})$
to $\ln(1-\delta+E_1/C_{1 \bs 1})$, representing growth rate for the model.
Storage effects are manifest by this function, because it is flat on the left and 
upward sloping on the right, so highly variable horizontal-axis inputs to the function
can translate into higher average vertical-axis values than can concentrated
horizontal-axis inputs of the same mean; e.g., the histograms labeled ``left'' on
E,F. Red dots indicate the mean values of
each histogram. The average vertical-axis values of the two red dots on each panel
equal the GWR of species $i$ for the model (value displayed numerically).

We now use Fig. \ref{fig:beta_fecundidities_explanatory} to explain ATAs. Coexistence is possible for 
right-tail association (panel F, positive GWR) but impossible for left-tail association (D,
negative GWR). Coexistence is possible for the right-tail case essentially
because the points on the ``left'' region in C produce highly variable
annual growth rates; the same effect does not occur for the left-tail case because 
points in the ``left'' region on panel A are concentrated on line $L_1$ because of the
strong left-tail association of $E_i$ and $C_{i \bs i}$ on that panel. 
Association, or lack thereof, in right tails makes essentially no difference 
to growth rates because $E_i/C_{i \bs i}$ cannot vary much. Overall, ATA effects 
are positive for the right-tail case because strong association between 
$E_i$ and $C_{i \bs i}$ is concentrated in the right tails, where there 
is little potential for storage effects; and so disassociation between
$E_i$ and $C_{i \bs i}$ in the left tails, where there is much potential for
storage effects, can be stronger. For the actual quantification of ATA effects,
a comparison is made to the case of symmetric association 
(panels B, E), as a benchmark. What is assessed by our quantification of ATA
effects is essentially the degree to which patterns of disassociations between 
$E$ and $C$, which can cause storage effects, align with the portions of 
the $E,C$ distribution for which the potential for storage effects is highest,
with larger ATA effects indicating better alignment, such as in panels C, F
of Fig. \ref{fig:beta_fecundidities_explanatory}.

\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/PostHocExplanatoryFig_betaFecund.jpg}
\caption{A detailed look at ATA effects for the beta fecundities lottery model.
The figure gives a pictorial representation of the calculation of  
$\E \left[ \ln \left( 1-\delta+\frac{E_i}{C_{i \bs i}} \right) \right]$ (A, C, D, F) and 
$\E \left[ \ln \left( 1-\delta+\frac{E_i^{||}}{C_{i \bs i}^{||}} \right) \right]$ (B, E)
for the model, considering both left- (A, D) and right-tail (C, F) associations
between $E_i$ and $C_{i \bs i}$. ATA effects are the difference of these two values.
The figure also illustrates how ATAs reflect the degree of alignment between where 
in the $EC$ distributions associations are allocated and which parts of that distribution
have high potential for storage effects. See SI section X2
%***DAN: fix this link when you put the text into the paper
for details. We used model parameters $\eta_1 = 1$, $\eta_2 = 1.2$, and $\delta=0.6$
for this example.}\label{fig:beta_fecundidities_explanatory}
\end{figure}
